Concurrent Programming with CXL

### 2020-06-16 book/paper.tex
Fixed author
Latest

\title{Concurrent Programming with CXL}
\author{RVR and possibly others}

@@

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes concurrently read and update shared variables and
synchronize with one another makes programs more complicated than
programs where only one thing happens at a time.  Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential program is mostly \emph{deterministic}.
\index{determinism}
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
The output of running concurrent programs depends on how the various
processes are interleaved.  Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\index{Heisenbug}
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically}
\index{atomicity}
because there is no other activity
interfering with their execution.  Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other instructions
may update memory locations as a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \texttt{count} and
an array \texttt{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \texttt{self}.
It increments \texttt{count} and sets \texttt{done[self]} to \texttt{True}.
Method \texttt{main} waits for the processes to finish by polling
both \texttt{done} flags.
After that, it verifies that the value of \texttt{count} equals 2.  If not,
it reports the actual value of \texttt{count}.
The program spawns three processes.
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.

@@

CXL programs manipulate CXL values.
CXL values are recursively defined:
they include booleans (\texttt{False} and \texttt{True}),
integers (but not floating point numbers),
strings (enclosed by double quotes),
sets of CXL values,
and dictionaries that map CXL values to other CXL values.
%
Another type of CXL value is the \emph{atom}.  It is essentially
just a name.  An atom is denoted using a period followed by the
name.  For example, \texttt{.main} is an atom.

@@

Instead of bits and bytes, a CXL machine manipulates CXL values.
A CXL machine has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CXL machine instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CXL machine has just one memory location containing
an CXL value.
\item Processes:  Any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
\index{name tag}
a program counter, a stack of CXL values,
and a single mutable general purpose \emph{register} that contains a CXL value.
\end{itemize}

@@

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
CXL values are unbounded trees.
Both the memory and the register of a process always contain
a dictionary that maps atoms to CXL values.  We call this a \emph{directory}.
A directory represents the state of a collection of variables named by the atoms.
%
Because directories are CXL values themselves,
directories can be organized into a tree.
Each node in a directory tree is then identified
by a sequence of atoms, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address} of a CXL value, and it is relative to the
``root'' of the directory tree.

@@

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A counter is a very simple example of
a data structure, but as we have seen it too requires multiple instructions.
A more involved one would be access to a binary tree.
Adding a node to a binary tree, or re-balancing a tree, often requires
multiple operations.  Maintaining ``consistency'' is certainly much easier
(although not necessarily impossible) if during this time no other
process also tries to access the binary tree.
An implementation of a data structure that can be safely accessed by multiple
processes (or threads) is called \emph{thread-safe}.

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CXL code.
Here \texttt{@cs} is a \emph{label}, identifying
a location in the CXL machine code.  The first thing we need to
ensure is that there cannot be two processes at the critical section.
This property is called \texttt{mutual exclusion}.
\index{mutual exclusion}
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

CXL in fact supports this.
It has an operator \texttt{atLabel $L$}, where $L$
is the atom containing the name of the label (in this case, \texttt{.cs}).
The operator returns a bag (multiset) of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.
Method \texttt{atLabel} only exists for specification purposes---do not
use it in normal code.
The assertion also makes use of the \texttt{nametag()} operator that
returns the name tag of the current process.
If you run the code through CXL, the assertion should fail because
there is no code yet for entering and exiting the critical section.

>>>>

\title{Concurrent Programming with CXL}
\author{Robbert van Renesse}

@@

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes concurrently read and update shared variables
\index{shared variable}
and synchronize with one another makes programs more complicated than
programs where only one thing happens at a time.  Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential program is mostly \emph{deterministic}.
\index{determinism}
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
The output of running concurrent programs depends on how the various
processes are interleaved.  Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\index{Heisenbug}
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically}
\index{atomicity}
because there is no other activity
interfering with their execution.  Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other instructions
may update memory locations as a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes
\index{process}
running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \texttt{count} and
an array \texttt{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \texttt{self}.
It increments \texttt{count} and sets \texttt{done[self]} to \texttt{True}.
Method \texttt{main} waits for the processes to finish by polling
both \texttt{done} flags.
After that, it verifies that the value of \texttt{count} equals 2.  If not,
it reports the actual value of \texttt{count}.
The program spawns three processes.
\index{process}
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.

@@

CXL programs manipulate CXL values.
CXL values are recursively defined:
they include booleans (\texttt{False} and \texttt{True}),
integers (but not floating point numbers),
strings (enclosed by double quotes),
sets of CXL values,
and dictionaries
\index{dictionary}
that map CXL values to other CXL values.
%
Another type of CXL value is the \emph{atom}.
\index{atom}
It is essentially
just a name.  An atom is denoted using a period followed by the
name.  For example, \texttt{.main} is an atom.

@@

Instead of bits and bytes, a CXL machine manipulates CXL values.
A CXL machine has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CXL machine instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CXL machine has just one memory location containing
an CXL value.
\item Processes:  Any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
\index{name tag}
a program counter,
\index{program counter}
a stack of CXL values,
and a single mutable general purpose \emph{register}
\index{register}
that contains a CXL value.
\end{itemize}

@@

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
CXL values are unbounded trees.
Both the memory and the register of a process always contain
a dictionary that maps atoms to CXL values.  We call this a \emph{directory}.
\index{directory}
A directory represents the state of a collection of variables named by the atoms.
%
Because directories are CXL values themselves,
directories can be organized into a tree.
Each node in a directory tree is then identified
by a sequence of atoms, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address} of a CXL value, and it is relative to the
``root'' of the directory tree.

@@

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A counter is a very simple example of
a data structure, but as we have seen it too requires multiple instructions.
A more involved one would be access to a binary tree.
Adding a node to a binary tree, or re-balancing a tree, often requires
multiple operations.  Maintaining ``consistency'' is certainly much easier
(although not necessarily impossible) if during this time no other
process also tries to access the binary tree.
An implementation of a data structure that can be safely accessed by multiple
processes (or threads) is called \emph{thread-safe}.
\index{thread safety}
\index{thread}

A critical section is often modeled as processes in an infinite loop
entering and exiting the critical section.
Figure~\ref{fig:csbarebones} shows the CXL code.
Here \texttt{@cs} is a \emph{label},
\index{label}
identifying a location in the CXL machine code.  The first thing we need to
ensure is that there cannot be two processes at the critical section.
This property is called \texttt{mutual exclusion}.
\index{mutual exclusion}
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.

CXL in fact supports this.
It has an operator \texttt{atLabel $L$},
\index{atLabel operator}
where $L$
is the atom containing the name of the label (in this case, \texttt{.cs}).
The operator returns a bag (multiset) of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.
Method \texttt{atLabel} only exists for specification purposes---do not
use it in normal code.
The assertion also makes use of the \texttt{nametag()} operator
\index{nametag operator}
that returns the name tag of the current process.
If you run the code through CXL, the assertion should fail because
there is no code yet for entering and exiting the critical section.
